---
title: "Session 5 Practice: Errors/error messages"
author: "Kenneth Tay"
date: "Oct 16, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

**Note: There are often multiple ways to answer each question.**  

For the following questions, explain what went wrong and how the program can be fixed.

Load these packages before starting the problems:
```{r warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
```

1. We want to add `x` and `y` together to get the value of 24:
```{r error=TRUE}
x <- 14
y <- "10"
x + y
```


2. We want to add `1+2` and divide it by `3+4`:
```{r error=TRUE}
((1+2)/(3+4)))
```


For the rest of the questions, we will use the `mtcars` dataset:
```{r}
data(mtcars)
```

3. We want to save the vector of numbers 1, 2, ..., L-1 into the variable `x`, where L is the number of columns in `mtcars`.
```{r}
x <- 1:ncol(mtcars)-1
x
```


4. We want to make a scatterplot of `mpg` vs. `wt`:
```{r error=TRUE}
ggplot(data = mtcars) +
    geom_point(y = mpg, x = wt)
```


5. We want to make a histogram of `mpg`:
```{r eval=FALSE}
ggplot(data = mtcars)
    + geom_histogram(aes(x = mpg))
```

```{r echo=FALSE, error=TRUE}
ggplot(data = mtcars)
    + geom_histogram(aes(x = mpg))
```


6. We want to make a boxplot of `mpg` for each value of `cyl`, overlay it with points, and add a title to the plot:
```{r eval=FALSE}
ggplot(data = mtcars, aes(x = cyl, y = mpg)) +
    geom_boxplot() +
    geom_point()
    labs(title = "Plot of mpg vs. cyl")
```

```{r echo=FALSE, error=TRUE}
ggplot(data = mtcars, aes(x = cyl, y = mpg)) +
    geom_boxplot() +
    geom_point()
    labs(title = "Plot of mpg vs. cyl")
```


7. We want a scatterplot of `qsec` vs. `wt`, but we want all the points to be colored blue:
```{r}
ggplot(data = mtcars, aes(y = qsec, x = wt)) +
    geom_point(aes(col = "blue"))
```


8. We want to create a new column which is `miles per quart` and display the first 3 rows:
```{r error=TRUE}
mtcars %>% mutate(miles per quart = mpg / 4) %>% head(n = 3)
```


9. We want to compute the mean `mpg` for each value of `gear`:
```{r error=TRUE}
mtcars %>% group_by(gear) %>% summarize(mean = mean)
```


10. We want to compute the maximum `hp` and `disp` in the dataset:
```{r}
mtcars %>% summarize(max = max(hp, disp))
```
