---
title: "Analysis of Student Alcohol Consumption"
author: "put your name"
date: "10/26/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

This is an analysis of how student alcohol consumption affects their academic scores. The data were obtained in a survey of students math and portuguese language courses in secondary school (dataset is available on [Kaggle](https://www.kaggle.com/uciml/student-alcohol-consumption).

We limit our analysis to the impact of alchol on math scores. While the dataset contains many interesting social, gender and study information (see [here](https://www.kaggle.com/uciml/student-alcohol-consumption) for more details on the columns), we will only look at a few columns that are of interest to us.

## Data import and processing

Library imports:
```{r}
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
```

Data import:
```{r}
original_df <- read_csv("student-mat.csv")
```

There are 395 students in the dataset, with 33 variables. For this analysis, the outcome that we are interested in understanding is G3 (final grade), which is numeric from 0 to 20. We want to understand how this is affected by a student's gender and alcohol consumption. The code below selects the relevant columns for our analysis:

```{r}
df <- original_df %>% select(sex, Dalc, Walc, score = G3)
kable(head(df))
```

## Differences in scores between males and females

First, we want to test if there is a difference in the distribution of scores between males and females. Let's look at this in a histogram:
```{r}
ggplot(data = df, mapping = aes(x = score)) +
    geom_histogram(aes(fill = sex), bins = 21) +
    facet_grid(sex ~ .)
```

For both sexes, we have a bimodal distribution: 1 mode at 0, and another around 10.

<!---------------
This is the place to insert code on testing difference between males and females
---------------->

## Relationship between scores and alcohol consumption

We are given 2 columns related to alchol consumption: `Dalc`, for consumption in the day, and `Walc`, for consumption at night. Both of these variables are numbers from 1 to 5. Note that these are not numeric variables: 1 is less than 2, but a value of 1 does not indicate half the consumption of 2. Strictly speaking, `Dalc` and `Walc` are **ordinal variables**: categorical variables with some inherent ordering of categories.

For this analysis, we will just treat them as numeric variables for simplicity. (We need more sophisticated tools to model ordinal variables correctly.)

Let's combine the 2 variables into one variable for alcohol by adding them together:
```{r}
df_alc <- df %>% mutate(alc = Dalc + Walc)
```

Scatterplot of `score` vs. `df_alc`:
```{r}
ggplot(data = df_alc, mapping = aes(x = alc, y = score)) + 
    geom_point(position = "jitter")
```

Let's fit a linear model of `score` vs. `alc`. Expectation: Increasing use of alcohol could have a negative effect on score.

<!---------------
This is the place to insert code on modeling linear relationship between score and alc
---------------->

## Modeling scores as a function of sex and alc

A person's gender could affect the relationship between scores and alcohol. Expectation: Regardless of gender, increasing use of alcohol could have a negative effect on score.

```{r}
ggplot(data = df_alc, mapping = aes(x = alc, y = score, col = sex)) + 
    geom_point(position = "jitter") +
    facet_wrap(~ sex)
```

<!---------------
This is the place to insert code on modeling score vs. alc and sex
---------------->
