---
title: "Session 6 Practice: Importing, transforming and plotting data"
author: "Kenneth Tay"
date: "Oct 18, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

**Note: There are often multiple ways to answer each question.**  

1. Download `nba_free_throws.csv` from https://github.com/kjytay/misc/tree/master/data. (Right click on `nba_free_throws.csv` and select "Save Link As..."). Import this dataset into R as the variable `df`. Are there columns which need their format changed?  

2. Select just the rows from the 2015-2016 regular season, remove the column `play` and save the result in `df2`.  

All questions from here are about `df2`.  

3. Display the top 10 players who took the most free throws.   

4. Free throw percentage is defined as the percentage of shots taken which were made. Display the top 10 players with the highest free throw percentages. (Hint: modify Qn 3 code to take two summaries.)

5. The highest free throw percentages are so high because these players didn't take many shots. Display the top 10 players with the highest free throw percentages **among only the players who took at least 100 free throws**. (Hint: Filter Qn 4 code at an appropriate step.)  

6. Save the summary table of shots taken, shots made and free throw percentage by player in `summary_df` (only players who took at least 100 free throws). Using `summary_df`, make a scatterplot of free throw percentage vs. free throws taken. Set the `alpha` value of the points to `0.5`, and draw a blue dashed horizontal line to show the mean free throw percentage across these players. (Hint: For the horizontal line, use `geom_abline`.)  

7. Which game (in `df2`) had the most number of free throws? Save the rows in `df2` from that game in `df3`.

8. Make a bar plot showing the number of free throws each player took in this game. Add `coord_flip()` as a layer to the plot so that the bars are horizontal. (Bonus: Can you sort the bars such that the longest ones go on top? The `forcats` package will be helpful, as will the last example of Section 15.4 of R4DS.)

The following code joins data from `summary_df` to `df3` and saves it as df4:
```{r eval=FALSE}
df4 <- df3 %>% left_join(summary_df, by = "player")
```

9. Modify your bar plot in Qn 8 so that the fill of the bars is equal to the player's free throw percentage. Add the layer `scale_fill_distiller(palette = "RdYlGn", direction = 1)` to give your bars some appropriate colors. Why are some bars grey?  


10. Using `tidyr`'s `separate` function, separate the `game` column in `df2` to a `home` column which has the name of the home team, and an `away` column which has the name of the away team. (See Section 12.4 of R4DS for details on the `separate` function.)  

