---
title: "Analysis of Student Alcohol Consumption"
author: "put your name"
date: "10/26/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

This is an analysis of how student alcohol consumption affects their academic scores. The data were obtained in a survey of students math and portuguese language courses in secondary school (dataset is available on [Kaggle](https://www.kaggle.com/uciml/student-alcohol-consumption).

We limit our analysis to the impact of alchol on math scores. While the dataset contains many interesting social, gender and study information (see [here](https://www.kaggle.com/uciml/student-alcohol-consumption) for more details on the columns), we will only look at a few columns that are of interest to us.

## Data import and processing

Library imports:
```{r}
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
```

Data import:
```{r}
original_df <- read_csv("student-mat.csv")
```

There are 395 students in the dataset, with 33 variables. For this analysis, the outcome that we are interested in understanding is G3 (final grade), which is numeric from 0 to 20. We want to understand how this is affected by a student's gender and alcohol consumption. The code below selects the relevant columns for our analysis:

```{r}
df <- original_df %>% select(sex, Dalc, Walc, score = G3)
kable(head(df))
```

## Differences in scores between males and females

First, we want to test if there is a difference in the distribution of scores between males and females. Let's look at this in a histogram:
```{r}
ggplot(data = df, mapping = aes(x = score)) +
    geom_histogram(aes(fill = sex), bins = 21) +
    facet_grid(sex ~ .)
```

For both sexes, we have a bimodal distribution: 1 mode at 0, and another around 10.

Let's compute the mean score for each of the sexes:
```{r}
df %>% group_by(sex) %>%
    summarize(mean_score = mean(score))
```

Test if the difference in mean scores for the sexes is significant or not with the $t$-test:
```{r}
m_scores <- (df %>% filter(sex == "M"))$score
f_scores <- (df %>% filter(sex == "F"))$score
t.test(m_scores, f_scores, alternative = "two.sided")
```

The $p$-value for this test is around 0.04, which is fairly low. We would probably reject the null hypothesis in favor of the alternative hypothesis.

Test if the distribution of male scores is significantly different from the distribution of female scores with the Kolmogorov-Smirnov test:
```{r}
ks.test(m_scores, f_scores, alternative = "two.sided")
```

The $p$-value for this test is around 0.15, so we don't have enough evidence to reject the null hypothesis (i.e. the data we have could have reasonably come from the distribution under the null hypothesis).

## Relationship between scores and alcohol consumption

We are given 2 columns related to alchol consumption: `Dalc`, for consumption in the day, and `Walc`, for consumption at night. Both of these variables are numbers from 1 to 5. Note that these are not numeric variables: 1 is less than 2, but a value of 1 does not indicate half the consumption of 2. Strictly speaking, `Dalc` and `Walc` are **ordinal variables**: categorical variables with some inherent ordering of categories.

For this analysis, we will just treat them as numeric variables for simplicity. (We need more sophisticated tools to model ordinal variables correctly.)

Let's combine the 2 variables into one variable for alcohol by adding them together:
```{r}
df_alc <- df %>% mutate(alc = Dalc + Walc)
```

Scatterplot of `score` vs. `df_alc`:
```{r}
ggplot(data = df_alc, mapping = aes(x = alc, y = score)) + 
    geom_point(position = "jitter")
```

Let's fit a linear model of `score` vs. `alc`. Expectation: Increasing use of alcohol could have a negative effect on score.

```{r}
lm(score ~ alc, data = df_alc)
```

Get more information on the linear fit with `summary`:
```{r}
fit <- lm(score ~ alc, data = df_alc)
summary(fit)
```

From the summary, it looks like alcohol consumption is uncorrelated with score.

Plot the linear fit along with the scatterplot:
```{r}
ggplot(data = df_alc, mapping = aes(x = alc, y = score)) + 
    geom_point(position = "jitter") +
    geom_smooth(method = "lm")
```

## Modeling scores as a function of sex and alc

A person's gender could affect the relationship between scores and alcohol. Expectation: Regardless of gender, increasing use of alcohol could have a negative effect on score.

```{r}
ggplot(data = df_alc, mapping = aes(x = alc, y = score, col = sex)) + 
    geom_point(position = "jitter") +
    facet_wrap(~ sex)
```

First, let's fit the additive model:
```{r}
fit <- lm(score ~ alc + sex, data = df_alc)
summary(fit)
```

The coefficient on `alc` is positive, suggesting that there is a positive relationship between alcohol and score! The p-value for the coefficient is small, so we shouldn't read too much into this positive relationship.

Next, let's fit the model with interactions:
```{r}
fit <- lm(score ~ alc * sex, data = df_alc)
summary(fit)
```

We can also draw the linear regression fits with the scatterplot:
```{r}
ggplot(data = df_alc, mapping = aes(x = alc, y = score, col = sex)) + 
    geom_point(position = "jitter") +
    geom_smooth(method = "lm", se = FALSE) +
    facet_wrap(~ sex)
```

## Optional: Fitting on new data

The following code creates a new dataset, and predicts the scores for the students in this dataset using the model with interaction effects:
```{r}
new_df <- data.frame(sex = c("F", "F", "M", "M"), 
                     alc = c(1, 5, 3, 7))

predict(fit, new_df)
```


## Optional: Drawing the linear fit for an additive model

I couldn't find an easy way to draw the linear fits for an additive model in the facetted scatterplot. (If you find a way, let me know!)

This is a workaround:
```{r}
# get fit coefficients
fit <- lm(score ~ alc + sex, data = df_alc)
coefs <- fit$coefficients

# add coefficients to dataset
df_alc$slope <- coefs[2]
df_alc$intercept <- ifelse(df_alc$sex == "F", coefs[1], coefs[1] + coefs[3])

# plot
ggplot(data = df_alc, mapping = aes(x = alc, y = score, col = sex)) + 
    geom_point(position = "jitter") +
    geom_abline(aes(slope = slope, intercept = intercept)) +
    facet_wrap(~ sex)
```
