---
title: "Session 1 Practice: Basic R objects"
author: "Kenneth Tay"
date: "Oct 2, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

**Note: There are often multiple ways to answer each question.**  

1. Create a vector of integers from -5 to 10 (inclusive) and assign it to the variable `x`.  

2. What is the sum of the elements of `x`? (This involves a function that we did not discuss in class. Google is your friend!)  

3. Create a vector which consists of the first 10 multiples of 3 (i.e. 3, 6, ..., 30) and assign it to the variable `y`.  

4. What is the result of `x + y`? Why does R return this result?  

5. What is the result of `"1" + "2"`? Why does R return this result?  

6. What is the result of `z <- c(1, 2, "3")`? Why does R return this result?  

7. Create a matrix using `M <- matrix(1, 4, 5)`. What is the result, and why does R return this result?  

8. Find the sum of the 3rd row of `M`.  

9. Create the following list:
```{r}
person <- list(name = "John Doe",
               age = 26,
               classes = c("ENG", "MAT", "SCI", "SPA", "MUS"))
```

What is the result of `person$classes[2]`? Why does R return this result?  

10. What code can I use to find out how many classes John Doe took?  
